---
title: "EDA"
format: html
editor_options: 
  chunk_output_type: console
---

## Introduction

Access to safe drinking water is an important public health concern. This project examines a water quality dataset containing measurements of several chemical and physical characteristics of water samples. The goal is to determine whether these measurements can be used to predict whether a sample is safe for drinking.

The response variable for this analysis is `Potability`, which indicates whether a water sample is safe to drink. A value of **1** represents potable water, while a value of **0** represents water that is not potable.

The predictor variables available in this dataset include:

- **ph** – pH value of the water.
- **Hardness** – concentration of dissolved calcium and magnesium.
- **Solids** – total dissolved solids (ppm).
- **Chloramines** – concentration of chloramines (ppm).
- **Sulfate** – sulfate concentration (mg/L).
- **Conductivity** – electrical conductivity of the water.
- **Organic_carbon** – organic carbon concentration (ppm).
- **Trihalomethanes** – concentration of trihalomethanes (μg/L).
- **Turbidity** – measure of water clarity.

The purpose of this exploratory data analysis (EDA) is to better understand the structure of the data, identify missing values, summarize the distributions of the variables, and explore relationships between the predictor variables and water potability. These results will guide the development of predictive models in the modeling portion of this project.

## Data

Load the packages we'll use in this section:

```{r}
# Load packages
library(tidyverse)
library(GGally)
```

Read in our data:

```{r}
# Import the data
water_data <- read_csv("water_potability.csv")

# take a look
glimpse(water_data)
```

`Potability` is a classification variable, so lets make it a factor rather than keeping it numeric. 

```{r}
water_data <- water_data |>
  mutate(Potability = as.factor(Potability))
```

We should check the dimensions of our dataset.

```{r}
# Dimensions of the data
dim(water_data)
```

Now lets check for missingness.

```{r}
# Missing values
colSums(is.na(water_data))
```

Looks like the `ph`, `Sulfate`, and `Trihalomethanes` variables have 491, 781, and 162 missing values, respectively. These missing values will be addressed during the modeling stage. Since the purpose of this document is exploratory, no imputation is performed here.

## Distribution of the Response Variable

Lets check the how our response variable is distributed.

```{r}
ggplot(water_data, aes(x = Potability)) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = "count",                   # adds our counts on top of the bars
            aes(label = after_stat(count)), 
            vjust = -0.5) +
  labs(x = "Potability",
       y = "Count",
       title = "Distribution of Water Potability") +
  theme_minimal()
```

## Summary Statistics

To begin the exploratory analysis, summary statistics were calculated separately for potable and non-potable water samples. Comparing the means and standard deviations provides an initial indication of whether the predictor variables differ between the two response groups.

```{r}
water_data_summary <- water_data |>
  group_by(Potability) |>
  summarize(
    across(
      where(is.numeric),
      list(
        Mean = ~mean(.x, na.rm = TRUE), # remove NA values
        SD = ~sd(.x, na.rm = TRUE)
      )
    )
  )

water_data_summary
```

This summary indicates that the predictor means are very similar for potable and non-potable water samples. In nearly every case, the difference in group means is small relative to the variability within each group, suggesting that no individual predictor strongly distinguishes the two classes.

## Pairs Plot

This `ggpairs` output provides a comprehensive summary of the relationships among the predictor variables.

```{r message = FALSE warning = FALSE}
ggpairs(
  water_data,
  columns = 1:9, # leave out column 10 (potability)
  mapping = aes(color = Potability, alpha = 0.3)
  )
```

The diagonal density plots indicate that most predictor variables have unimodal distributions with substantial overlap between potable and non-potable water samples. The scatterplots do not indicate any strong linear relationships between predictor variables, and the correlation coefficients confirm that most pairwise correlations are relatively weak (|r| < 0.37). Overall, the exploratory analysis suggests that no single predictor is sufficient to classify water potability on its own. Instead, combining several predictors within a machine learning model is likely to provide greater predictive performance.

## Discussion

The exploratory analysis suggests that predicting water potability is likely to be a challenging classification problem. The predictor variables do not exhibit strong separation between potable and non-potable samples individually, and the relationships among predictors are generally weak. Consequently, it is unlikely that any single variable will provide accurate predictions on its own. Instead, machine learning methods that combine information from multiple predictors may be better suited for identifying patterns associated with water potability. The next section therefore develops and compares classification tree and random forest models using the tidymodels framework

[Click here for the Modeling Page](Modeling.html)
